---
title: "State of Arctic models"
output: html_document
---
  
```{r packages and data, include=FALSE}
require(knitr)
require(tidyverse)
require(lubridate)
require(nlme) #for mixed effects models
require(broom) #for nested modelling
require(modelr)
require(gridExtra) #for grid.arrange
require(broom) #for do and tidy
require(ggfortify) #for autoplot
require(pander) #for pander table

#set directories
wd = "D:/Box Sync/Arctic/CONNECT/Paper_5_state_of_arctic/Analysis/Intermediate"

#Industries
industries <- c("Cruise_tourism", "Domestic_tourism", "International_tourism", "Population", "Shipping_distance", "Hunters", "Fishing", "Oilandgas_wells")

#a function to change the plot sizes
subchunkify <- function(g, fig_height=7, fig_width=5) {
  g_deparsed <- paste0(deparse(
    function() {g}
  ), collapse = '')
  
  sub_chunk <- paste0("
  `","``{r sub_chunk_", floor(runif(1) * 10000), ", fig.height=", fig_height, ", fig.width=", fig_width, ", echo=FALSE}",
  "\n(", 
    g_deparsed
    , ")()",
  "\n`","``
  ")
  
  cat(knitr::knit(text = knitr::knit_expand(text = sub_chunk), quiet = TRUE))
}

```

```{r merge data, include=FALSE}
#load and merge files
alldat <- lapply(industries, function(i) {
  print(paste0("loading ", i))
    
    dat <- read.csv(paste0(wd, "/", i, "_long.csv"), header=TRUE, fileEncoding = "UTF-8-BOM", stringsAsFactors = FALSE)
    dat$value_raw <- dat$value
    dat$subIndustry <- i
    if(i %in% c("Cruise_tourism", "Domestic_tourism", "International_tourism")){
        dat$Industry <- "Tourism"
    } else {
        dat$Industry <- i
    }
    
      if(i == "Shipping_distance"){ 
        dat <- dat %>% filter(Country!="Total") %>% 
                  group_by(Country, Region, Industry, subIndustry, year) %>% 
                  summarise(value=sum(value, na.rm=TRUE)) %>% #sum shipping distance across all types of shipping
                  mutate(Metric="nautical miles sailed all shipcats") %>% 
                  mutate(value_raw=value)
        dat$value[dat$value==0] <- 1 #add small constant to make growth rate calculation possible
        
      } 
    
      if(i == "Fishing"){ 
        dat <- dat %>% filter(Region!="Total") 
      } 
      

      if(i == "Cruise_tourism"){ 
      #Murmansk cruise tourism statistics use a different metric than elsewhere, number of ship calls as opposed to passengers.
      #We convert ship calls to an estimation of passenger numbers by multiplying the number of calls by the average vessel capacity of ships calling in 2016 (437.4 passengers).
      #adjust murmansk
      dat$value[dat$Region=="Murmansk"] <-  dat$value[dat$Region=="Murmansk"]*437.4
      dat[dat$Region=="Murmansk", "Metric"] <- "Passengers"
      #drop other metrics
      dat <- dat %>% filter(Metric=="Passengers")
      } 
      
      if(i == "Hunters") { 
        #drop other types of hunting
        dat <- dat %>% filter(Metric %in% c("All hunting/trapping/fishing licenses", "Number of hunters who paid game management fees", "Registered hunters", "Paid hunting permits", "Sports hunters"))
        #drop extra column
        dat <- dat[, which(!names(dat)=="Region_original")]
      }
      
      if(i == "Oilandgas_wells") {
        #remove duplicated data: drop Region="all" - this is a sum of onshore and offshore; and purpose="all" - this is sum of exploration & development
        dat <- dat %>% filter(Region!="All" & Purpose!="All"|is.na(Purpose) & Metric!="All") %>% #have to explicitly select NAs to get Alaska etc
              group_by(Country, Region, Metric, year) %>% #the metric column tells us whether exploration or dev
              summarise(value_raw=sum(value_raw, na.rm=TRUE)) %>% #sum the different types of wells (injection, wildcat etc)
              mutate(subIndustry=paste("Oilandgas", Metric, sep="_")) #split Industry into exploration and development      
              dat$Metric <- "n_wells"
              dat$Industry <- "OilandGas"

        #change value to cumsum
        dat <- dat %>% group_by(Country,Region,subIndustry) %>% 
                arrange(year) %>%
                mutate(value=cumsum(value_raw)) %>% #cumulative sum of n wells
                mutate(value=replace(value, value==0, NA)) #replace zeros with NA
      }
      
    return(data.frame(dat))

})
alldat <- bind_rows(alldat)
#alldat$value[alldat$value==0] <- NA #replace zeros with NA, otherwise anngrowthfun returns Inf
alldat$Country[alldat$Country %in% c("Svalbard & Jan Mayen", "Svalbard and Jan Mayen")] <- "Svalbard"
alldat <- alldat %>% mutate_at(c("Country", "Region", "Industry", "subIndustry"), funs(factor)) %>% #convert to factors
                    filter(!is.na(value)) #drop NAs
write_excel_csv(alldat, "All_industries_long.csv") #for fileEncoding = "UTF-8-BOM"

```

We calculate the annual growth rate in X over start years (y0) value (Xy0)  
annual growth rate = {log10(Xy1+zif) - log10(Xy0+zif)}*(1/nyrs)

in the output table, annual growth is assigned to final year (y1) - so annual growth between 1999 and 2003, is assigned to 2003.  
Where data is missing for several years, the annual growth rate is calculated from the two time points we have data for. Note that we don't include the no-data points in the model i.e. perform no interpolation for missing data.

### Exceptions:
#### Shipping distance: 
In some parts of the Arctic there is no shipping in certain years (e.g. Davis Sea in 2012). I added a small constant (zif) to these numbers, because log10(0) evaluates to -Inf, which can't be modelled.

#### Oil and gas:
We calculate annual growth in oil and gas from the cumulative number of wells drilled across time, as activity is intermittent (lots of years with no wells drilled). This gives growth rates of zero in years with no activity, rather than growth rates of Inf. This can be loosely taken as a measure of the growth (both rise and fall) in the industry across time.


```{r growthrates}
  #calculate annual growth
zif=0.0000001

alldat <- read.csv("All_industries_long.csv", header=TRUE, fileEncoding = "UTF-8-BOM")

anngrowthdf <- alldat %>%  filter(!is.na(value)) %>% #drop years missing data
    group_by(Industry, subIndustry, Country, Region, Metric)  %>%
    arrange(year) %>% #arrange by ascending years
    mutate(nyr = year - lag(year, 1)) %>% #number of years between subsequent datapoints
    mutate(anngrowth = {log(value+zif) -log(lag(value,1)+zif)}*(1/nyr))  %>% #annual growth between those years
    drop_na() %>% droplevels()

write_excel_csv(anngrowthdf, "All_industries_annualgrowth.csv") #for fileEncoding = "UTF-8-BOM"
  
  head(anngrowthdf)
 
```

Next we calculate the intensity of each industry, using the mean value across 2012-2017.  
we use a spread of years rather than a single year to account for interannual variation and missing data.

```{r intensities} 

intensitydf <- alldat %>%  filter(!is.na(value) & !Industry %in% c("OilandGas")) %>% #drop years missing data
  filter(year %in% 2012:2017) %>% 
  group_by(Country, Region, Industry, subIndustry, Metric)  %>%
  summarise(intensity=mean(value)) 
```

We calculate the intensity for oil and gas diffently, as the total number of wells drilled between 1960 and 2017.

```{r fixintensity1}
oilintensity <- alldat %>%  filter(!Industry %in% c("OilandGas")) %>% 
  filter(year %in% 1960:2017) %>% 
  group_by(Country, Industry, subIndustry, Region, Metric)  %>%
  summarise(intensity=sum(value_raw, na.rm=TRUE))

#replace oil and gas values with these new values
intensitydf <- bind_rows(intensitydf, oilintensity)
intensitydf <- intensitydf %>% drop_na() %>% droplevels()
write_excel_csv(intensitydf, "All_industries_intensity.csv") #for fileEncoding = "UTF-8-BOM"

```

Murmansk cruise tourism statistics use a different metric than elsewhere, number of ship calls as opposed to passengers.
We convert ship calls to an estimation of passenger numbers by multiplying the number of calls by the average vessel capacity of ships calling in 2016 (437.4 passengers). This gives higher than expected cruise passengers - same intensity as Troms...
```{r, include=FALSE}
alldat %>% filter(subIndustry=="Cruise_tourism" & Region %in% c("Murmansk", "Troms", "Finnmark", "Arkhangel'sk") & year==2016) %>% select(Region, value)
```

Let's plot the data.  

## Annual growth rates
I removed one datapoint for Shipping_distance in Davis Strait, where growth rate in 2002-2013 was 5180% (up from no shipping in 2012 to 518nm in 2013)  
Oil and gas development shows a high growth rate in 2005 for the Barent's Sea, due to the start of production (up from 1 well in 2004 to 9 in 2005)  

```{r plot growthrates, echo=FALSE, warning=FALSE}

levels(anngrowthdf$subIndustry)

anngrowthdf %>% filter(Industry=="Tourism" & year %in% c(1997:2017)) %>%
  ggplot(aes(x=year, y=anngrowth, color=Country)) +
  geom_line() +
  facet_wrap(~ subIndustry, scales="free_y") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

anngrowthdf %>% filter(!Industry %in% c("Tourism", "Fishing", "Shipping_distance") & year %in% c(1997:2017)) %>%
  ggplot(aes(x=year, y=anngrowth, color=Country)) +
  geom_line() +
  facet_wrap(~ subIndustry, scales="free_y") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

anngrowthdf %>% filter(Industry=="Fishing" & year %in% c(1997:2017)) %>%
  ggplot(aes(x=year, y=anngrowth, color=Region)) +
  geom_line() +
  ggtitle("Fishing catch") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

anngrowthdf %>% filter(Industry=="Shipping_distance" & year %in% c(2012:2017)) %>%
  ggplot(aes(x=year, y=anngrowth, color=Region)) +
  geom_line() +
  ggtitle("Shipping distance") +
  theme_minimal() +
  ylim(-1, 1) +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

```

## Intensities  
The plots for Domestic and International tourism and Hunters should be interpreted with caution - different countries use different metrics.  

#### Domestic & International tourism
Canada, Alaska, Russia = number of visitors    
All other countries = guest nights 

#### Hunters
Alaska = All hunting/trapping/fishing licenses  
Greenland = Sports hunters,  
Norway = Registered hunters,  
Finland & Sweden = Paid hunting permits,  
No data for Russia, Faroes, Iceland, Canada.  

```{r plot intensities, echo=FALSE, warning=FALSE}
intensitydf %>% filter(Industry=="Tourism") %>%
ggplot(aes(x=Country, y=intensity)) +
  geom_col() +
  facet_wrap(~ subIndustry, scales="free_y") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

intensitydf %>% filter(!Industry %in% c("Tourism", "Fishing", "Shipping_distance")) %>%
ggplot(aes(x=Country, y=intensity)) +
  geom_col() +
  facet_wrap(~ subIndustry, scales="free_y") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

intensitydf %>% filter(Industry=="Fishing") %>%
ggplot(aes(x=Region, y=intensity)) +
  geom_col() +
  ggtitle("Fishing catch") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

intensitydf %>% filter(Industry=="Shipping_distance") %>%
ggplot(aes(x=Region, y=intensity)) +
  geom_col() +
  ggtitle("Shipping distance") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

```  

## Models
It seems like the inter-annual variation is high, and no trends are immediately obvious in any industry.  
Is the annual growth significantly different from zero? Let's model it.  

First we run a very simple model of growth ~ year. We might expect different trends in each country (influenced by policy, marketing, resource availability, economic & demographic conditions etc), so let's also run a linear model with country base model of growth ~ year + country. Finally, we explore an interaction between year and country and see if it outperforms the base model.
simple  = growth ~ year
base    = growth ~ year + country
interaction = growth ~ year * country

```{r models, echo=FALSE, warning=FALSE}
#set up data
#anngrowthdf$Region2 <- as.character(anngrowthdf$Country)
anngrowthdf$Region2 <- as.character(anngrowthdf$Region)
anngrowthdf$Region2[anngrowthdf$subIndustry %in% c("Fishing", "Shipping_distance")] <- as.character(anngrowthdf$Region[anngrowthdf$subIndustry %in% c("Fishing", "Shipping_distance")])
anngrowthdf$Region2 <- as.factor(anngrowthdf$Region2)

#run the models across the industries
dfmods <- anngrowthdf %>% 
  filter(!subIndustry %in% c("Fishing", "Shipping_distance")) %>%
  filter(year>1997 & year<2018) %>%
  group_by(subIndustry) %>%
  do(basemod = lm(anngrowth ~ I(year-2017) + relevel(Region2, ref="Iceland"), data=.), 
     intmod = lm(anngrowth ~ I(year-2017) * relevel(Region2, ref="Iceland"), data=.),
     simplemod = lm(anngrowth ~ I(year-2017), data=.))  
#run the models across the industries fishing
dfmods_fish <- anngrowthdf %>% 
  filter(subIndustry=="Fishing") %>%
  filter(year>1997 & year<2018) %>%
  group_by(subIndustry) %>%
  do(basemod = lm(anngrowth ~ I(year-2017) + relevel(Region2, ref="Barents Sea"), data=.), 
     intmod = lm(anngrowth ~ I(year-2017) * relevel(Region2, ref="Barents Sea"), data=.),
     simplemod = lm(anngrowth ~ I(year-2017), data=.))  
#run the models across the industries shipping
dfmods_ship <- anngrowthdf %>% 
  filter(subIndustry=="Shipping_distance") %>%
  filter(year>1997 & year<2018) %>%
  group_by(subIndustry) %>%
  do(basemod =lm(anngrowth ~ I(year-2017) + relevel(Region2, ref="NorwayEEZ"), data=.), 
     intmod = lm(anngrowth ~ I(year-2017) * relevel(Region2, ref="NorwayEEZ"), data=.),
     simplemod = lm(anngrowth ~ I(year-2017), data=.)) 

#check if the interaction model performs better
overallmodelfit <- bind_rows(glance(dfmods, simplemod), glance(dfmods, basemod), glance(dfmods, intmod)) %>%
                   bind_cols(data.frame(model = rep(c("simple", "base", "interaction"), each=nrow(dfmods)))) %>%
                  arrange(subIndustry)
overallmodelfit_fish <- bind_rows(glance(dfmods_fish, simplemod), glance(dfmods_fish, basemod), glance(dfmods_fish, intmod)) %>%
                   bind_cols(data.frame(model = rep(c("simple", "base", "interaction"), each=nrow(dfmods_fish))))
overallmodelfit_ship <- bind_rows(glance(dfmods_ship, simplemod), glance(dfmods_ship, basemod), glance(dfmods_ship, intmod)) %>%
                   bind_cols(data.frame(model = rep(c("simple", "base", "interaction"), each=nrow(dfmods_ship))))
overallmodelfit <- bind_rows(overallmodelfit, overallmodelfit_fish, overallmodelfit_ship)

knitr::kable(overallmodelfit, caption="Model selection")
#write.table(overallmodelfit, paste0(dirname(wd), "/Model/Allindustries_modelfit.csv"))

#print model summary
#lapply(dfmods$basemod, summary)
#lapply(dfmods$intmod, summary)
#modelsummary_base <- dfmods %>% tidy(basemod)
#modelsummary_int <- dfmods %>% tidy(intmod)

#set up the prediction values
#dfmodpreds_base <- dfmods %>% augment(basemod)
#dfmodpreds_int <- dfmods %>% augment(intmod)

```

### Plot growth against year for the entire Arctic.

```{r models_simple, echo=FALSE, warning=FALSE}
for(i in levels(anngrowthdf$subIndustry)) {
#run the models across the industries
  currdat <- anngrowthdf %>% 
    filter(year>1997 & year<2018) %>%
    filter(subIndustry==i)
  currmod <- lm(anngrowth ~ I(year-2017), currdat)
  print(summary(currmod))
  currdf <- augment(currmod)
    names(currdf)[2] <- c("year")
      #plot fitted
      p1 <- currdf %>%
        ggplot(aes(year, anngrowth)) +
        geom_point() +
        geom_abline(intercept=coefficients(currmod)[1], slope=coefficients(currmod)[2]) +
        ggtitle(sprintf("Linear trend %s", i)) +
        theme_minimal() +
        theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
    p2 <- currdf %>% 
       ggplot(aes(year, .resid)) +
        geom_point() +
        ggtitle("Remaining pattern") +
        theme_minimal() +
        theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
    grid.arrange(p1, p2, ncol=2)
}


```

### Plot growth against year, including region as an additive term.

```{r models_base, echo=FALSE}
printfun <- function(data){
        #pander(data[[2]][[1]])
        print(summary(data[[2]][[1]]))
        print(autoplot(data[[2]][[1]], which=c(1:3)))
        print(autoplot(data[[2]][[1]], which=c(5))  + theme(axis.text.x  = element_text(angle=90, vjust=0.5)))
}

for(i in levels(anngrowthdf$subIndustry)) {
  print(i)
  #plot model
  if(i %in% c("Fishing", "Shipping_distance")) {
  
    #subset data
    if(i=="Fishing"){
      currdf <- dfmods_fish 
      #print models
      printfun(currdf)
      #predict models
      currdf <- dfmods_fish %>% augment(basemod)
        names(currdf)[1:4] <- c("subIndustry", "anngrowth", "year", "Region2")  
    } else if (i =="Shipping_distance"){
      currdf <- dfmods_ship  
      #print models
      printfun(currdf)
      #predict models
      currdf <- dfmods_ship %>% augment(basemod)
        names(currdf)[1:4] <- c("subIndustry", "anngrowth", "year", "Region2")  
    }
  #plot fitted
  p1a <- currdf %>%
    filter(Region2 %in% levels(droplevels(currdf$Region2))[1:6]) %>%
    ggplot(aes(year, anngrowth)) +
      geom_point() +
      geom_line(aes(year, .fitted), inherit.aes = FALSE) +
      ggtitle("Linear trend") +
      facet_wrap(~ Region2, scales="free_y", ncol=1, strip.position = "top") +
      theme_minimal() +
      theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
  p1b <- currdf %>%
    filter(Region2 %in% levels(droplevels(currdf$Region2))[7:12]) %>%
    ggplot(aes(year, anngrowth)) +
      geom_point() +
      geom_line(aes(year, .fitted), inherit.aes = FALSE) +
      facet_wrap(~ Region2, scales="free_y", ncol=1, strip.position = "top") +
      theme_minimal() +
      theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
  p1c <- currdf %>%
    filter(Region2 %in% levels(droplevels(currdf$Region2))[13:length(levels(droplevels(currdf$Region2)))]) %>%
    ggplot(aes(year, anngrowth)) +
      geom_point() +
      geom_line(aes(year, .fitted), inherit.aes = FALSE) +
      facet_wrap(~ Region2, scales="free_y", ncol=1, strip.position = "top") +
      theme_minimal() +
      theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
   p2a <- currdf %>% 
     filter(Region2 %in% levels(droplevels(currdf$Region2))[1:6]) %>%
    ggplot(aes(year, .resid)) +
      geom_line() +
      ggtitle("Remaining pattern") +
      facet_wrap(~ Region2, scales="free_y", ncol=1, strip.position = "top") +
      theme_minimal() +
      theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
  p2b <- currdf %>% 
    filter(Region2 %in% levels(droplevels(currdf$Region2))[7:12]) %>%
    ggplot(aes(year, .resid)) +
      geom_line() +
      facet_wrap(~ Region2, scales="free_y", ncol=1, strip.position = "top") +
      theme_minimal() +
      theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
  p2c <- currdf %>% 
    filter(Region2 %in% levels(droplevels(currdf$Region2))[13:length(levels(droplevels(currdf$Region2)))]) %>%
    ggplot(aes(year, .resid)) +
      geom_line() +
      facet_wrap(~ Region2, scales="free_y", ncol=1, strip.position = "top") +
      theme_minimal() +
      theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
  grid.arrange(p1a, p2a, ncol=2)
  grid.arrange(p1b, p2b, ncol=2)
  grid.arrange(p1c, p2c, ncol=2)
  
  } else {
    #print models
    currdf <- dfmods  %>% filter(subIndustry==i)
    printfun(currdf)
    #predict models
    currdf <- currdf %>% augment(basemod)
    names(currdf)[1:4] <- c("subIndustry", "anngrowth", "year", "Region2")
    #plot fitted
    p1 <- currdf %>%
      ggplot(aes(year, anngrowth)) +
      geom_point() +
      geom_line(aes(year, .fitted), inherit.aes = FALSE) +
      ggtitle("Linear trend") +
      facet_wrap(~ Region2, scales="free_y", ncol=1, strip.position = "right") +
      theme_minimal() +
      theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
  p2 <- currdf %>% 
     ggplot(aes(year, .resid)) +
      geom_line() +
      ggtitle("Remaining pattern") +
      facet_wrap(~ Region2, scales="free_y", ncol=1, strip.position = "right") +
      theme_minimal() +
      theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
  grid.arrange(p1, p2, ncol=2)
  }
}
```

NEXT STEPS:
drop outliers


```{r eval=FALSE, include=FALSE}
by_industry <- anngrowthdf %>%
    filter(year>1997 & year<=2018) %>%
    group_by(Industry) %>% 
    nest()  
#starting with a simple general linear model
# no interaction, so we model the same growth rate in each country
#simple_model <- function(df){
#    if(df$subIndustry %in% c("Fishing", "Shipping_distance")){
 #   lm(anngrowth ~ year + Region, data=df)
 #   } else {
 #   lm(anngrowth ~ year + Country, data=df)
 #   }
#} 
interaction_model <- function(df){
    if(df$subIndustry %in% c("Fishing", "Shipping_distance")){
    glm(anngrowth ~ year*Region, data=df)
    } else {
    glm(anngrowth ~ year*Country, data=df)
    }
}

  #predictions
  grid <- testdf %>% 
    data_grid(year, Country) %>%
    gather_predictions(testm1, testm2)
  #residuals
  testdf <- testdf %>% 
    gather_residuals(testm1, testm2)
  #plot predictions
  ggplot(testdf, aes(year, anngrowth, color=Country)) +
    geom_point() +
    geom_line(data=grid, aes(x=year, y=pred, color=Country), inherit.aes = FALSE) +
    facet_wrap(~ model)
  #plot residuals
  ggplot(testdf, aes(year, resid, color=Country)) +
    geom_point() +
    facet_grid(Country ~ model)
  



testm1 <- simple_model(testdf)
testm2 <- interaction_model(testdf)
summary(testm1)
summary(testm2)
grid <- testdf %>% 
  data_grid(year, Country) %>%
  gather_predictions(testm1, testm2)

ggplot(testdf, aes(year, anngrowth, color=Country)) +
  geom_point() +
  geom_line(data=grid, aes(x=year, y=pred, color=Country), inherit.aes = FALSE) +
  facet_wrap(~ model)

testdf <- testdf %>% 
  gather_residuals(testm1, testm2)

ggplot(testdf, aes(year, resid, color=Country)) +
  geom_point() +
  facet_grid(Country ~ model)


#We want to know if there are global trends in Arctic industries, so let's include country as a random effect.
#mod3 <- lme(anngrowth ~ year + random=~1|Country, data=anngrowthdf, method='REML') #plus country as random effect
#mod4 <- gls(anngrowth ~ year, data=anngrowthdf, method='REML') #fixed effects
#anova(mod3, mod4)

```


