
---
title: "State of Arctic intensity adjustment"
output: html_document
---
  
```{r packages and data, include=FALSE}

require(sf)
require(tidyverse)
wd <- "D:/Box Sync/Arctic/CONNECT/Paper_5_state_of_arctic/Analysis/Model/"
```

### Intensity of activities
We calculate the intensity of each industry using the mean value across 2012-2017.  
we use a spread of years rather than a single year to account for interannual variation and missing data.
As tourism is increasing, we use the 2016 values

```{r intensities, include=FALSE} 
alldat <- read.csv("D:/Box Sync/Arctic/CONNECT/Paper_5_state_of_arctic/Analysis/Model/All_industries_long.csv", header=TRUE, fileEncoding = "UTF-8-BOM")

intensitydf <- alldat %>%  filter(!is.na(value) & !Industry %in% c("OilandGas", "Mining", "Tourism")) %>% #drop years missing data
  filter(year %in% 2012:2017) %>% 
  group_by(Country, Region, Industry, subIndustry, Metric)  %>%
  summarise(intensity=mean(value)) 
```

We calculate the intensity for oil and gas diffently, as the total number of wells drilled between 1960 and 2017, and for mining as the total number of mines active at any time during 1960 and 2017.

```{r fixintensity1, include=FALSE}
oilintensity <- alldat %>%  filter(Industry %in% c("OilandGas")) %>% 
  filter(year %in% 1960:2017) %>% 
  group_by(Country, Industry, subIndustry, Region, Metric)  %>%
  summarise(intensity=sum(value_raw, na.rm=TRUE))

miningintensity <- read.csv("D:/Box Sync/Arctic/Data/Landuse/Mining/Ojas_mining_database_2018/processed/Mining_listofMines_withdates.csv")
miningintensity <- miningintensity %>% 
    filter(Status!="Exploration" & is.na(Stop)|Stop>=1960) %>%
    group_by(Country, Region) %>% 
    summarise(intensity=n_distinct(X)) %>%
    mutate(Industry="Mining",
           subIndustry="Mining", 
           Metric="Operational_mines")

tourismintensity <- alldat %>%  filter(!is.na(value) & Industry %in% c("Tourism")) %>% #drop years missing data
  filter(year==2016) %>% 
  group_by(Country, Region, Industry, subIndustry, Metric)  %>%
  summarise(intensity=value) 


#replace oil and gas values with these new values
intensitydf <- bind_rows(intensitydf, oilintensity, miningintensity, tourismintensity)
intensitydf <- intensitydf %>% drop_na() %>% droplevels()
write_excel_csv(intensitydf, paste0(wd, "All_industries_intensity.csv")) #for fileEncoding = "UTF-8-BOM"

```

Murmansk cruise tourism statistics use a different metric than elsewhere, number of ship calls as opposed to passengers.
We convert ship calls to an estimation of passenger numbers by multiplying the number of calls by the average vessel capacity of ships calling in 2016 (437.4 passengers). This makes no difference to the zscore, but puts the intensity on the same scale as other regions. This gives higher than expected cruise passengers - same intensity as Troms...

```{r, include=FALSE}
alldat %>% filter(subIndustry=="Cruise_tourism" & Region %in% c("Murmansk", "Troms", "Finnmark", "Arkhangel'sk") & year==2016) %>% select(Region, value)
```

The plots for Domestic and International tourism and Hunters should be interpreted with caution - different countries use different metrics.  

#### Domestic & International tourism
Canada, Alaska, Russia = number of visitors    
All other countries = guest nights 

#### Hunters
Alaska = All hunting/trapping/fishing licenses  
Greenland = Sports hunters,  
Norway = Registered hunters,  
Finland & Sweden = Paid hunting permits,  
No data for Russia, Faroes, Iceland, Canada.  


```{r loadshps, include=FALSE}
area_tourism <- read_sf("D:/Box Sync/Arctic/Data/Boundaries/Arctic_circle/State_of_Arctic/SOA_borders_for_tourism2.shp") 
  st_geometry(area_tourism) <- NULL
  area_tourism[area_tourism$Region=="Northern Ostrobothnia", "Region"] <- "North Ostrobothnia"
  area_tourism[area_tourism$Region=="Svalbard & Jan Mayen", "Region"] <- "Svalbard and Jan Mayen"
  area_tourism <- area_tourism %>% group_by(Region) %>% summarise(Area_km = sum(Area_km))
area_cruise <- read.csv("D:/Box Sync/Arctic/Data/Boundaries/Arctic_circle/State_of_Arctic/Arctic_coastlines.csv", header=TRUE, stringsAsFactors = FALSE)
area_popn <- read_sf("D:/Box Sync/Arctic/Data/Boundaries/Arctic_circle/State_of_Arctic/SOA_borders_for_popn2.shp")
  st_geometry(area_popn) <- NULL
  area_popn <- area_popn %>% group_by(Region) %>% summarise(Area_km = sum(Area_km))
  area_popn[area_popn$Region=="Northern Ostrobothnia", "Region"] <- "North Ostrobothnia"
  area_popn[area_popn$Region=="Svalbard & Jan Mayen", "Region"] <- "Svalbard and Jan Mayen"
  area_popn <- area_popn %>% bind_rows(area_tourism[area_tourism$Region=="Alaska",])
area_eez <- read_sf("D:/Box Sync/Arctic/Data/Boundaries/Arctic_circle/State_of_Arctic/EEZ_plus_highseas_noland.shp")
  st_geometry(area_eez) <- NULL
  area_eez <- area_eez %>% group_by(Region) %>% summarise(Area_km = sum(Area_km))
  area_eez[area_eez$Region=="OpenDavisStraight" & !is.na(area_eez$Region), "Region"] <- "OpenDavisStrait"
area_fishing <- read_sf("D:/Box Sync/Arctic/Data/Boundaries/Arctic_circle/State_of_Arctic/SOA_borders_for_fishing.shp")
  st_geometry(area_fishing) <- NULL
  area_fishing[area_fishing$Region=="Fareo Plateau", "Region"] <- "Faroe Plateau"
  area_fishing <- area_fishing %>% group_by(Region) %>% summarise(Area_km = sum(Area_km))
area_mining <- read_sf("D:/Box Sync/Arctic/Data/Boundaries/Arctic_circle/State_of_Arctic/SOA_borders_for_mining.shp") 
  st_geometry(area_mining) <- NULL
  area_mining <- area_mining %>% group_by(Region) %>% summarise(Area_km = sum(Area_km))
area_mining2 <- read_sf("D:/Box Sync/Arctic/Data/Boundaries/Arctic_circle/State_of_Arctic/SOA_borders_for_mining2.shp") 
  st_geometry(area_mining2) <- NULL
  area_mining2 <- area_mining2 %>% group_by(Region) %>% summarise(Area_km = sum(Area_km))

intensitydf <- read.csv(paste0(wd, "All_industries_intensity.csv"), header=TRUE, fileEncoding = "UTF-8-BOM", stringsAsFactors = FALSE) #for fileEncoding = "UTF-8-BOM"
```

```{r adjustintensity, include=FALSE}
#Hunters
hunters <- intensitydf %>% filter(Industry=="Hunters") %>% 
            left_join(area_mining, by="Region") %>% 
            mutate(adj_intensity = intensity/Area_km)

#population
popn <- intensitydf %>% filter(Industry=="Population") %>%
          left_join(area_popn, by="Region") %>% 
          mutate(adj_intensity = intensity/Area_km)

#Mining
mining <- intensitydf %>% filter(Industry=="Mining") %>%
      left_join(area_mining2, by="Region") %>% 
      mutate(adj_intensity = 1000*intensity/Area_km)

#Oil and gas
oilandgas <- intensitydf %>% filter(Industry=="OilandGas") %>%
      mutate(Region=replace(Region, Region=="Chukchi Sea", "Northern Bering - Chukchi Seas")) %>%
      mutate(Region=replace(Region, Country=="Greenland" & Region=="Offshore", "Greenland Sea")) %>%
      mutate(Region=replace(Region, Country=="Greenland" & Region=="Onshore", "Greenland")) %>%
      mutate(Region=replace(Region, Country=="United Kingdom", "UKEEZ")) %>%
      mutate(Region=replace(Region, Region=="Labrador offshore", "Newfoundland-Labrador Shelf")) %>%
      mutate(Region=replace(Region, Region=="NWT", "Northwest Territories")) %>%
     left_join(area_fishing, by="Region") %>% 
     left_join(area_mining, by="Region") %>% 
     left_join(area_eez, by="Region") %>% 
      mutate(Area_km = rowSums(.[7:9], na.rm=TRUE)) %>%
      select(-c(Area_km.x, Area_km.y)) %>%
      mutate(adj_intensity = 1000*intensity/Area_km)

#Shipping_distance
shipping <- intensitydf %>% filter(Industry=="Shipping_distance") %>%
  left_join(area_eez, by="Region") %>% 
  mutate(adj_intensity = intensity/Area_km)

#Fishing
fishing <- intensitydf %>% filter(Industry=="Fishing") %>%
  left_join(area_fishing, by="Region") %>% 
  mutate(adj_intensity = intensity/Area_km)

#Tourism
tourism <- intensitydf %>% filter(subIndustry %in% c("International_tourism","Domestic_tourism")) %>%
  left_join(area_tourism, by="Region") %>% 
  mutate(adj_intensity = intensity/Area_km)

#cruise tourism by coastline
cruise <- intensitydf %>% filter(subIndustry=="Cruise_tourism") %>%
  left_join(area_cruise, by="Region") %>% 
  mutate(adj_intensity = intensity/Coastline_km)

#Save
intensitydf_new <- bind_rows(hunters, popn, mining, oilandgas, shipping, fishing, tourism, cruise)
write.csv(intensitydf_new, paste0(wd, "All_industries_intensity_perkm.csv"), row.names=TRUE)
```

# PLOT
```{r plot, include=TRUE}
#load data adjusted by area
#intensitydf <- read.csv(paste0(wd, "All_industries_intensity_perkm.csv"),  header=TRUE, fileEncoding = "UTF-8-BOM", stringsAsFactors = FALSE)

intensitydf_new %>% filter(Industry=="Tourism") %>%
  ggplot(aes(x=Country, y=intensity)) +
  geom_col() +
  facet_wrap(~ subIndustry, scales="free_y") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

intensitydf_new %>% filter(!Industry %in% c("Tourism", "Fishing", "Shipping_distance")) %>%
  ggplot(aes(x=Country, y=intensity)) +
  geom_col() +
  facet_wrap(~ subIndustry, scales="free_y") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

intensitydf_new %>% filter(Industry=="Fishing") %>%
  ggplot(aes(x=Region, y=intensity)) +
  geom_col() +
  ggtitle("Fishing catch") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

intensitydf_new %>% filter(Industry=="Shipping_distance") %>%
  ggplot(aes(x=Region, y=intensity)) +
  geom_col() +
  ggtitle("Shipping distance") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
```